---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Baldur

The goal of Baldur is to shine light on your proteomics data.

## Installation
You can install the development version of `baldur` from this github or the stable version from CRAN.
Importantly, you first need to follow the instructions for installing `rstan` [https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started](https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started).
Then you can install `baldur` as accordingly:

You can install the development version of `baldur` from this github:
```{r}
devtools::install_github('PhilipBerg/baldur', build_vignettes = T)
# or
# install.package('baldur')
```

## Example

Please see the vignettes for an examples `vignette('baldur_yeast_tutorial')` and `vignette('baldur_ups_tutorial')`.
